---
title: "Machine Learning 1"
author: "Saumya Ranjan (A15483401)"
date: "10/21/2021"
output: html_document
---

First is clustering methods 

# K means clustering 

The Function in base R to do K means clustering is called "kmeans()"

First make up some data where we know what the answer should be:


```{r}
tmp <- c(rnorm(30, -3), rnorm(30, 3))
x <- cbind(x=tmp, y=rev(tmp))
plot(x)
```


Question: can we use kmeans() to cluster this data setting k 2 and nstart to 20

```{r}
km <- kmeans(x, centers=2, nstart = 20)

km
```

Question: How many point are in each cluster? 
```{r}
km$size
```
30 points in each cluster 

Question: What ‘component’ of your result cluster assignment/membership?      
```{r}
km$cluster
```
Question: What component of your result object details cluster center? 

```{r}
km$centers
```


Queston: Plot x colored by kmeans cluster assignment and add cluster centers as blue points 
```{r}
plot(x, col=km$cluster)
points(km$centers, col="blue", pch=15, cex=2)
```


# H clust

A big limitation w kmeans is that we have to  tell it K (the number of clusters we want)

Analyze the same data with hclust

```{r}
hc <- hclust( dist(x) )
hc
```

There is a plot method for hclust result objects. Let's see it

```{r}
plot(hc)
```

To get our cluster membership vector we have to do a bit more work. We have to  cut the tree where we think it makes more sense. For this we use 'cutree()' function 

```{r}
cutree(hc, h=6)
```
You can also call cutree() setting k= the number of clusters or groups you want 

```{r}
grps <- cutree(hc, k=2)
```

Make our results plot
```{r}
plot(x, col=grps)
```


## PCA of UK food data
```{r}
url <- "https://tinyurl.com/UK-foods"
x <- read.csv(url)
```

Q1. How many rows and columns are in your new data frame named x? What R functions could you use to answer this questions?
```{r}
nrow(x)
ncol(x)
```

Checking your data
```{r}
head(x)
tail(x)
```
Want to have 4 columns. Right now the row names are set to be one column of its own 
```{r}
url <- "https://tinyurl.com/UK-foods"
x <- read.csv(url, row.names=1)
head(x)
```
Q2. Which approach to solving the ‘row-names problem’ mentioned above do you prefer and why? Is one approach more robust than another under certain circumstances?
-- the one where we change the row name in the beginning. If we use "x <- x[,-1]" then every time we run the code, a column gets deleted. 


Now we have the data looking good we want to explore it. We will use some conventional plots 
```{r}
barplot(as.matrix(x),col=rainbow(nrow(x)))
barplot(as.matrix(x), beside=T, col=rainbow(nrow(x)))
```
Q3: Changing what optional argument in the above barplot() function results in the following plot?
--changing the "beside" argument. If beside= true, the bars will be beside each other.


Q5: Generating all pairwise plots may help somewhat. Can you make sense of the following code and resulting figure? What does it mean if a given point lies on the diagonal for a given plot?
```{r}
pairs(x, col=rainbow(10), pch=16)
```
The function creates plots that pair 2 countries and plots the results in a scatter plot. If the line lies on the diagonal it means that the value for the 2 countries are the same; there is no variation between the 2 for that catagory. 

Q6. What is the main differences between N. Ireland and the other countries of the UK in terms of this data-set?
-- we can compare N Ireland to the rest of the UK by looking at how a category differs from the other countries. Like I can see in the bar graph that N Ireland eats less of what the "'light blue" group compared to the rest of the UK. But making comparisons are hard. 

##PCA to the rescue

```{r}
pca <- prcomp(t(x))
summary(pca)
```

Q7. Complete the code below to generate a plot of PC1 vs PC2. The second line adds text labels over the data points.
```{r}
# Plot PC1 vs PC2
plot(pca$x[,1], pca$x[,2], xlab="PC1", ylab="PC2", xlim=c(-270,500))
text(pca$x[,1], pca$x[,2], colnames(x), col=c("orange", "Red", "blue", "green"))
```


##Digging deeper (variable loadings)
```{r}
par(mar=c(10, 3, 0.35, 0))
barplot( pca$rotation[,1], las=2 )
```
 
 
 Q9: Generate a similar ‘loadings plot’ for PC2. What two food groups feature prominantely and what does PC2 maninly tell us about?
```{r}
par(mar=c(10, 3, 0.35, 0))
barplot( pca$rotation[,2], las=2 )
```
-- the two food groups that feature are fresh potatoes and soft drinks. It tells us that the variation for N Ireland (where most of the differences between N Ireland and the rest of the UK) lies in fresh potato and soft drink consumption 



```{r}
## The inbuilt biplot() can be useful for small datasets 
biplot(pca)
```


##2. PCA of RNA-seq data
```{r}
url2 <- "https://tinyurl.com/expression-CSV"
rna.data <- read.csv(url2, row.names=1)
head(rna.data)
```
Q10: How many genes and samples are in this data set?
-- There are 100 genes and 10 samples

```{r}
pca <- prcomp(t(rna.data), scale=TRUE)
 
## Simple un polished plot of pc1 and pc2
plot(pca$x[,1], pca$x[,2], xlab="PC1", ylab="PC2")
```
```{r}
summary(pca)
plot(pca, main="Quick scree plot")
```

Making the scree plot ourselves 
```{r}
## Variance captured per PC 
pca.var <- pca$sdev^2

## Percent variance
pca.var.per <- round(pca.var/sum(pca.var)*100, 1)
pca.var.per

barplot(pca.var.per, main="Scree Plot", 
        names.arg = paste0("PC", 1:10),
        xlab="Principal Component", ylab="Percent Variation")
```

Making the PCA plot look colorful 
```{r}
colvec <- colnames(rna.data)
colvec[grep("wt", colvec)] <- "red"
colvec[grep("ko", colvec)] <- "blue"

plot(pca$x[,1], pca$x[,2], col=colvec, pch=16,
     xlab=paste0("PC1 (", pca.var.per[1], "%)"),
     ylab=paste0("PC2 (", pca.var.per[2], "%)"))

text(pca$x[,1], pca$x[,2], labels = colnames(rna.data), pos=c(rep(4,5), rep(2,5)))
```


Using ggplot 
```{r}
library(ggplot2)

df <- as.data.frame(pca$x)

# Basic PCA1 vs PCA2 plot
ggplot(df) + 
  aes(PC1, PC2) + 
  geom_point()
```

Labeling WT and knockout samples 
```{r}
# Add a 'wt' and 'ko' "condition" column
df$samples <- colnames(rna.data) 
df$condition <- substr(colnames(rna.data),1,2)

p <- ggplot(df) + 
        aes(PC1, PC2, label=samples, col=condition) + 
        geom_label(show.legend = FALSE)
p
```

Adding themes and title+subtitles + percent var to axis labels 
```{r}
p + labs(title="PCA of RNASeq Data",
       subtitle = "PC1 clealy seperates wild-type from knock-out samples",
       x=paste0("PC1 (", pca.var.per[1], "%)"),
       y=paste0("PC2 (", pca.var.per[2], "%)"),
       caption="BIMM143 example data") +
     theme_bw()
```

